Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-11-27 inst/paper/main.tex
more bib

@@

R is a lazy functional programming language~\cite{ecoop12} with
dynamic features that allow to write object-oriented code.  Most data
types are vectorized. Values are constructed by the \c{c(...)}
function, e.g. \c{c("hi","ho")} creates a vector of two strings.  To
enable equational reasoning, R copies values accessible through
multiple variables when they are written to.  Values can be tagged by
user-defined attributes. For instance, one can attach the attribute
\c{dim} to the value \c{x<-c(1,2,3,4)} with \c{attr(x,"dim")<-c(2,2)}.
This causes arithmetic functions to treat \c x as a $2 \times 2$
matrix. Another attribute is \c{class} which can be bound to a list of
names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
\c{human}; classes are used for object-oriented dispatch.  Every R
linguistic construct is desugared to a function call, even control
flow statements, assignments, and bracketing. Furthermore, all
functions can be redefined in user code. This makes R both flexible
and challenging to compile~\cite{dls19}.  Arguments to user-defined
functions are bundled into thunks called \emph{promises}. Logically, a
promise combines an expression's code, its environment, and its value.
To access the value of a promise, one must force it. Forcing a promise
triggers evaluation and the computed value is captured for future
reference.

@@

@@

Originally, \eval was limited to interpreted languages, but with the
advent of just-in-time compilation this restriction was gradually
lifted.

Some languages have chosen to restrict \eval. The essential difference
between languages relates to scoping of variables that can be accessed
from code executed by \eval.  Java and Julia are among the more
restrictive languages. In Java, \eval must be synthesized as a
combination of dynamic loading and reflection.\footnote{While Java
does not have \eval built-in, one can take the input to \eval, wrap it
in a static method of an anonymous class, generate bytecode for that
class, invoke the Java class loader to dynamically install the new
code, and then use reflection to call the new method.} Julia provides
\eval more directly but limits its scope to top level variables and
methods. JavaScript supports both a, so-called, global \eval, and a
local \eval, where the latter has acces to local variables of the
function being executed (as well as variables of lexically enclosing
scopes). Finally, languages such as R allow \eval to even manipulate
variables defined in the caller environment of the function in which
\eval is invoked.

The difference between these design choices has an impact on the
ability of compilers to optimize, and of analysis tools to reason
about code. The impact of a top level \eval is that any global
variable may be modified, but currently running functions can continue
to execute unchanged. A local \eval, may change the values of local
variables, their types, or even their presence. This is a much more
invasive change, one that may invalidate currently executing functions
as it may change the results of common optimizations such as constant
folding or copy propagation. The designers of the Julia programming
language were aware of this difference and purposefully chose to
restrict \eval to the top level in order to preserve their ability to
optimize code~\cite{oopsla18a}. They even added a versioning mechanism
to function definitions to ensure that code added during an \eval
would not invalidate optimizations such as inlining.  R's ability to
affect any one of the callers of a function is even more invasive as
the impact of \eval is non-local. Such unfettered power is difficult
to deal with.

>>>>

This section provides a short introduction to R as it has a few
surprising features that impact the use of reflective features of the
language. We then look at the semantics of \eval in R and discuss some
design choices. Lastly, we put this paper in context of previous work.

@@

R is a lazy functional programming language~\cite{ecoop12} with
dynamic features that allow to write object-oriented code. Most data
types are vectorized. Values are constructed by the \c{c(...)}
function, e.g. \c{c("a","bc")} creates a vector of two strings. To
enable equational reasoning, R copies values accessible through
multiple variables when they are written to. Values can be tagged by
user-defined attributes. For instance, one can attach the attribute
\c{dim} to the value \c{x<-c(1,2,3,4)} with \c{attr(x,"dim")<-c(2,2)}.
This causes arithmetic functions to treat \c x as a $2 \times 2$
matrix. Another attribute is \c{class} which can be bound to a list of
names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
\c{human}; classes are used for object-oriented dispatch. Every R
linguistic construct is desugared to a function call, even control
flow statements, assignments, and bracketing. Furthermore, all
functions can be redefined in user code. This makes R both flexible
and challenging to compile~\cite{dls19}. Arguments to user-defined
functions are bundled into thunks called \emph{promises}. Logically, a
promise combines an expression's code, its environment, and its value.
To access the value of a promise, one must force it. Forcing a promise
triggers evaluation and the computed value is captured for future
reference.

@@

The expressive power of \eval depends on a number of choices made by
the programming language designers. The two axes that are the most
relevant here are:
\begin{itemize}
\item {\bf Scoping:} What environment does the string passed to \eval
  execute in? JavaScript and R evaluate the expression in the current
  scope. This means that local variables and parameters are exposed to
  \eval, breaking the abstraction boundary of the enclosing function.
  Julia takes a more restrictive approach by running \eval at the
  ``top level'', i.e. in the global environment, outside of any
  function's environment. JavaScript has a global \eval that behaves
  like Julia, it also has a ``strict mode'' in which \eval is allowed
  to access local variables, but fobidden from injecting new variables
  in the local environment. In Java, one can implement \eval but the
  expression will execute in new environment.\footnote{While \eval
  used to be available only in interpreted languages, the advent of
  just-in-time compilation gradually lifted this restriction. To
  implement \eval in Java requires wrapping its input in a static
  method of an anonymous class, generating bytecode for that class,
  invoking the class loader to install the code, and then using
  reflection to call the new method. }
\item {\bf Reflective API:} What other reflective operations are
  exposed to user code? JavaScript does provides few reflective
  functions (other than being able to enumerate the properties of an
  object and to access properties with strings), the string passed to
  \eval is thus limited to what can be expressed in JavaScript. R, on
  the other hand, has a rich reflective interface. For example, R
  allows user code to walk the call stack and arbitrarily add and
  delete local variables in any environment.
\end{itemize}

\noindent
These design choices impact our ability to reason about code. More
expressive power for \eval translates directly into additional
restriction for program analysis and transformation tools. The reason
Julia has restricts \eval to execute in top level environment is that
they wanted to preserve their compiler's ability to optimize
functions~\cite{oopsla18a}. What they wanted to prevent was the case
where \eval changed the type of local variables as this would require
recompilation of the method (or prevent optimization). Julia even adds
a versoning mechanism, called world age, to ensure that code added
during an \eval does not invalidate optimizations such as
inlining~\cite{oopsla20a}.

R's ability to affect any one of the callers of a function is even
more invasive as the impact of \eval is non-local. Such unfettered
power is difficult to deal with.

@@
